---
title: "Applying scSLIDE to large-scale single-cell data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scSLIDE_SEA_AD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This vignette will illustrate how to utilize the [BPCells](https://satijalab.org/seurat/articles/seurat5_bpcells_interaction_vignette) (developed by [Benjamin Parks](https://github.com/bnprks/BPCells)) functionality in Seurat to apply scSLIDE to large-scale single-cell data. We will also utilize [future R-package](https://cran.r-project.org/web/packages/future/index.html)'s parallelization to improve the runtime.We will be using a publicly available [Alzheimer's disease snRNA-seq dataset from the Allen Brain Institute](https://www.nature.com/articles/s41593-024-01774-5) for demonstration.


## Setup and Library Loading

First, we load all necessary libraries and set up the working environment. 

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(SeuratObject, lib = "/home/jiangl/R")
library(Seurat, lib = "/home/jiangl/R")
library(scSLIDE, lib = "/home/jiangl/R")
library(AzimuthAPI, lib = "/home/jiangl/R")
library(princurve, lib = "/home/jiangl/R")
library(ggplot2, lib = "/home/jiangl/R")
library(ggridges, lib = "/home/jiangl/R")
library(Scillus, lib = "/home/jiangl/R")
library(future, lib = "/home/jiangl/R")
library(knn.covertree, lib = "/home/jiangl/R")
library(destiny, lib = "/home/jiangl/R")
library(BPCells, lib = "/home/jiangl/R")

# set the future.globals.maxSize to 10GB
# options(future.globals.maxSize = 10 * 1000 * 1024^2)
# allocate 4 CPUs for parallelization
# plan(multicore, workers = 4)

# create a file name 
DATA_NAME = "SEA_AD_atlas_MTG"

```


```{r setup, eval=FALSE}
# some other packages that will be used in this vignette can be installed using the following commands:
# remotes::install_github("bnprks/BPCells/r")
# remotes::install_github("xmc811/Scillus", ref = "development")
# BiocManager::install("destiny")
# install.packages(c("princurve", "ggridges"))

library(SeuratObject)
library(Seurat)
library(scSLIDE)
library(AzimuthAPI)
library(princurve)
library(ggplot2)
library(ggridges)
library(Scillus)
library(future)
library(knn.covertree)
library(destiny)
library(BPCells)

# create a file name 
DATA_NAME = "SEA_AD_atlas_MTG"
```

## Generating on-disk Seurat object from .h5ad file

We will use the SEA-AD snRNA-seq data (~1.4m cells) from [Gabitto et al 2024 Nat. Neurosci.](https://www.nature.com/articles/s41593-024-01774-5) as an example for our workflow. This dataset can be downloaded from [CZI Cellxgene data portal ](https://cellxgene.cziscience.com/collections/1ca90a2d-2943-483d-b678-b809bf464c30). The are several datasets available at this link, and we will be using the data collected from patients' MTG (middle temporal gyrus) region.  
\
The original file is in .h5ad [AnnData](https://anndata.readthedocs.io/en/stable/) format. Users can load the whole data into R using packages like the [anndata R-package](https://cran.r-project.org/web/packages/anndata/readme/README.html), but this is often not memory-efficinet enough. Below, we will show you how to generate an on-disk Seurat object based on the .h5ad file, and apply our scSLIDE to this on-dick data in our following analyses.  
\
Note that before proceeding to the following workflow, if you want to annotate your single-cell .h5ad file with our Pan-Human Azimuth python tool, you can follow the simple instruction [provided here](https://satijalab.org/pan_human_azimuth/). 

<details><summary style="font-size: 20px;"><strong>Click here to see how to generate an on-disk Seurat object from .h5ad</strong></summary>

```{r message = FALSE, results = 'hide', warning = FALSE, eval = FALSE}
# path to the h5ad file (assuming it has been annotated by the Pan-Human Azimuth python tool)
file_path = "./SEA_AD_atlas_MTG_ANN.h5ad"

# the path to save the on-disk matrix object (from BPCells)
ondisk_path = paste0("./", DATA_NAME, "_ondisk/")

# load the count matrix with BPCells
ondisk.data <- open_matrix_anndata_hdf5(file_path, group = "/X")
# write this BPCells matrix to disk
write_matrix_dir(mat = ondisk.data, dir = ondisk_path)

# get necessary components from the original h5ad file
# feature names
feature_metadata = schard::h5ad2data.frame(file_path, 'var')
# metadata
metadata = schard::h5ad2data.frame(file_path, 'obs')
# get the pan-human Azimuth embeddings / umap
ann_embedding = t(schard::h5ad2Matrix(file_path,
                                      '/obsm/azimuth_embed'))

# now load the on-disk BPCells data 
ondisk.mat <- open_matrix_dir(dir = ondisk_path)

# rename the rownames/features using gene symbol instead of ENSG IDs
rownames(ondisk.mat) = feature_metadata$feature_name

# finally, create an on-disk Seurat object and add the metadata; This object links to the `ondisk.mat` BPCells matrix.
ondisk.object <- CreateSeuratObject(counts = ondisk.mat, meta.data = metadata)

# add low-dim objects to the Seurat object
# add ann_embeddings 
rownames(ann_embedding) = colnames(ondisk.object)
ondisk.object[['azimuth_embed']] = CreateDimReducObject(embeddings = ann_embedding, assay = "RNA", key = "Ann_")

##### Basic QC and preprocessing for the on-disk data
# nCount and nFeature
ondisk.object <- subset(ondisk.object, subset = nCount_RNA >= 1000 & nFeature_RNA >= 500)
# remove cells with discordant ANN_celltype_labels
ondisk.object <- subset(ondisk.object, subset = full_consistent_hierarchy == TRUE)
# remove cells with very low final_level prob
ondisk.object <- subset(ondisk.object, subset = final_level_softmax_prob >= 0.6)

# next we will remove cells from cell types with cell count less than 50
ct_table = table(ondisk.object$Subclass)
keep_celltype = names(ct_table)[ct_table >= 50]
ondisk.object <- subset(ondisk.object, subset = Subclass %in% keep_celltype)

# generate necessary metadata columns 
ondisk.object$Donor = as.character(ondisk.object$donor_id)
ondisk.object$dementia = ondisk.object$disease
# rename the disease 
ondisk.object$Disease = "HV"
ondisk.object$Disease[!ondisk.object$ADNC %in% c("Not AD", "Reference")] = "AD"
ondisk.object$Disease = factor(ondisk.object$Disease, levels = c("HV", "AD"))

# fix the 'Age' and 'Sex' meta-data
ondisk.object$Sex = ondisk.object$sex
# Note that those donors labelled with "80 year-old and over human stage" are actually >= 90 years old (cause it is inconsistent with 'Age at death'), so we need to manually fix it.
ondisk.object$Age = gsub(pattern = "-year-old human stage", replacement = "", ondisk.object$development_stage)
ondisk.object$Age[ondisk.object$development_stage == "80 year-old and over human stage"] = 90
ondisk.object$Age = as.numeric(ondisk.object$Age)

# save the object for later use
saveRDS(ondisk.object, file = paste0("Seurat_obj_1_OnDisk_", DATA_NAME, ".rds"))
rm(ondisk.object, ann_embedding, metadata, feature_metadata, ondisk.mat, ondisk.data)
gc()
```

<p style="text-align: center; margin-top: 20px;">
<a href="#" onclick="this.closest('details').open = false; return false;">Click here to collapse ▲</a>
</p>
</details>


## Loading on-disk data and preprocessings

Here, we will load the on-disk Seurat object into R, and process it just like a regular in-memory object with normalization and finding variable features.

```{r message = FALSE, results = 'hide', warning = FALSE, include = FALSE}
# load the Seurat on-disk object 
object = readRDS(paste0("/brahms/jiangl/test_ondisk_2025Apr01/Seurat_obj_1_OnDisk_", DATA_NAME, ".rds"))

# basic processing
object = NormalizeData(object)
object = FindVariableFeatures(object, nfeatures = 2000)
```


```{r message = FALSE, results = 'hide', warning = FALSE, eval=FALSE}
# load the Seurat on-disk object 
object = readRDS(paste0("./Seurat_obj_1_OnDisk_", DATA_NAME, ".rds"))

# basic processing
object = NormalizeData(object)
object = FindVariableFeatures(object, nfeatures = 2000)
```


## Prepare the Single-Cell Data for Sample-Level Analysis

To perform all necessary pre-processing steps, we provide a wrapper function called `PrepareSampleObject()`. This function prepares a single-cell Seurat object for downstream sample-level density analysis. It (optionally) identifies additional highly variable genes through correlation testing, creates training and landmark cell subsets via sketching methods, performs PLS learning to capture response-relevant variation, and constructs a weighted multi-modal nearest neighbor graph between landmark cells and the full dataset. The function outputs a prepared Seurat object containing the landmark assay and a cell-to-landmark WNN graph needed for subsequent sample object generation.  
\
Note that here we are specifying `max_core =` and `future.memory.per.core =` to utilize [future's multicore](https://cran.r-project.org/web/packages/future/index.html) parallelization functionality to reduce the runtime. We will require 4 cores and a total of 3000 Mb * 4 shared memory for this process. Note that "multicore" is not supported by R-studio.


```{r message = FALSE, results = 'hide', warning = FALSE}
object <- PrepareSampleObject(object = object,
                             assay = "RNA",
                             add.hvg = T,
                             group.by.CorTest = "Subclass",
                             Y = "Disease",
                             sketch.training = T,
                             group.by.Sketch = "Donor",
                             ncells.per.group = 3000,
                             ncells.landmark = 5000,
                             ncomp = 20,
                             pls.function = "cppls",
                             k.nn = 5,
                             name.reduction.1 = "azimuth_embed",
                             dims.reduction.1 = 1:128,
                             fix.wnn.weights = c(0.5, 0.5),
                             max_core = 4,
                             future.memory.per.core = 3000,
                             rm.training.assay = T,
                             verbose = F)
```


## Sample-Level Object Generation

Once a WNN graph is constructed, we can generate a sample-level density matrix. Given the asymmetric landmark-by-cell WNN graph, scSLIDE aggregates over cells to obtain a landmark-by-sample density matrix (m landmarks by n samples) by binarizing adjacency (neighbor = 1, else 0) and summing per sample. Each entry in this landmark-by-sample matrix measures the cell density near each landmark for a given donor.

The function `GenerateSampleObject()` automatically generates this sample-level object based on the WNN object specified by `nn.name =`. Additionally, it helps users search for donor-level metadata from the original single-cell object and append it to the sample-level object (optional). It also performs a Chi-squared normalization (based on Poisson distribution of a contingency table) for the count matrix (optional; standard log-normalization is also supported).

```{r message = FALSE, results = 'hide', warning = FALSE}
sample_obj = GenerateSampleObject(object = object,
                                  sketch.assay = "LANDMARK",
                                  nn.name = "weighted.nn",
                                  group.by = "Donor",
                                  rename.group.by = "Subclass",
                                  k.nn = 5,
                                  normalization.method = "ChiSquared",
                                  add.meta.data = TRUE,
                                  return.seurat = T,
                                  new_assay_name = "LMC"
                                  )

# we can filter out donors with cell count less than 200 (i.e., nCount_LMC = k.nn * cell_count < 1000)
sample_obj = subset(sample_obj, subset = nCount_LMC >= 1000)

# Center the data
sample_obj = ScaleData(sample_obj, features = rownames(sample_obj), do.scale = F, do.center = T)
```


## Diffusion Map Analysis

To better denoise the sample-level data, instead of performing standard PCA, we will perform Diffusion Map (a non-linear unsupervised dimensional reduction method originally proposed by [Haghverdi et al Bioinformatics](https://academic.oup.com/bioinformatics/article/31/18/2989/241305) to capture single-cell trajectories).

```{r message = FALSE, results = 'hide', warning = FALSE}
sample_obj = RunDiffusionMap(sample_obj, assay = "LMC", layer = "scale.data", features = rownames(sample_obj), ncomp = 5)
```

By visualizing the Diffusion Map components, we can see that there seems to be a progression trajectory along DC1 and DC2 jointly.

```{r fig.width=7, fig.height=6}
DimPlot(sample_obj, reduction = "DiffMap", group.by = "Disease") 
```

We append DC1-3 to the metadata and analyze their relationships with clinical variables.

```{r}
# Append DC1-3 to the metadata
sample_obj[['traj_DC1']] = sample_obj@reductions$DiffMap@cell.embeddings[, 1]
sample_obj[['traj_DC2']] = sample_obj@reductions$DiffMap@cell.embeddings[, 2]
sample_obj[['traj_DC3']] = sample_obj@reductions$DiffMap@cell.embeddings[, 3]
```


### Principal Curve Fitting

We can see that the first two DCs seem to represent a joint trajectory. We will fit a principal curve to them to infer a pseudo-trajectory curve.

```{r message = FALSE, results = 'hide', warning = FALSE, fig.width=7, fig.height=6}
princurve_res = princurve::principal_curve(sample_obj[['DiffMap']]@cell.embeddings[, 1:2],
                                           plot = F, maxit = 20, trace = T)
sample_obj[['traj_overall']] = princurve_res$lambda
DimPlot(sample_obj, reduction = "DiffMap", group.by = "Disease")  +
  geom_path(aes(x = princurve_res$s[princurve_res$ord, 1], y = princurve_res$s[princurve_res$ord, 2]))
```


The following plots show that the principal curve is related to the neuropathological continuous pseudo-progression score (CPS) provided by the original paper. The CPS values were obtained by aggregating different patient-level neuropathalogical measturements from their postmortem brains, and it is independent from their scRNA-seq data. 

```{r message = FALSE, results = 'hide', warning = FALSE, fig.width=7, fig.height=6}
mdata <- sample_obj@meta.data
mdata$bin_traj_overall <- cut(rank(mdata$traj_overall), breaks = 5, labels = paste0("traj-bin_", 1:5))
# 
p11 = ggplot(data = mdata) +
    geom_boxplot(aes(x = bin_traj_overall,
                     y = `Continuous Pseudo-progression Score`),
                 outliers = F) +
    geom_point(aes(x = bin_traj_overall,
                   y = `Continuous Pseudo-progression Score`,
                   col = Disease,
                   shape = Disease),
               position = position_jitter(width = 0.2, height = 0)) +
    ggtitle("Pseudo-time vs. CPS") +
    theme_bw() +
    theme(title = element_text(hjust = 0.5, size = 15),
          text = element_text(size = 13),
          axis.text.x = element_text(hjust = 1, vjust = 1, angle = 45),
          axis.text = element_text(size = 13)) +
    xlab("Pseudo-trajectory bins") +
    ylab("Continuous Pseudo-progression Score")

print(p11)
```

## Differential Expression Analysis Setup

Finally, we can perform differential expression (DE) analysis to extract differentially expressed genes (DEGs) that change along the pseudo-trajectory we inferred. 

First, we need to obtain the pseudobulked expression data for the original single-cell data:

```{r}
bulk_obj = AggregateExpression(object, assays = "RNA", group.by = c("Donor", "Subclass"), return.seurat = T)

# Append the metadata to pseudobulked object 
mdata = object@meta.data[, c("Donor", "Subclass")]
mdata = mdata[!duplicated(mdata), ]
rownames(mdata) = paste0(gsub(pattern = "_", replacement = "-", mdata$Donor),
                         "_",
                         gsub(pattern = "_", replacement = "-", mdata$Subclass))
bulk_obj = AddMetaData(bulk_obj, metadata  = mdata)

bulk_obj$nCount_RNA = colSums(x = bulk_obj, slot = "counts")  # nCount_RNA
bulk_obj$nFeature_RNA = colSums(x = LayerData(object = bulk_obj, layer = "counts") > 0)  # nFeature_RNA
```

### DE Analysis for different cell types

Then we run DE analysis for one cell type at a time. We will use "Microglia-PVM" as an example.

```{r}
sub_bulk_obj = subset(bulk_obj, subset = Subclass == "Microglia-PVM")

# Add metadata from sample-level object to pseudobulk object
new_mdata = merge(sub_bulk_obj@meta.data[, c("Donor"), drop = F],
                  sample_obj@meta.data[, c("Donor", "Age", "Sex", "Disease", "traj_overall")],
                  by = "Donor",
                  all.x = T,
                  sort = FALSE)
sub_bulk_obj = AddMetaData(sub_bulk_obj, metadata = new_mdata)
sub_bulk_obj$log_count = log10(sub_bulk_obj$nCount_RNA)
```

We can now run trajectory DE tests.

```{r message = FALSE, results = 'hide', warning = FALSE}
de_results = TrajDETest(sub_bulk_obj, assay = "RNA", layer = "counts",
                        traj.var = "traj_overall", latent.vars = c("Age", "Sex", "log_count"))

```

Now, for the DE results, we will identify the top up- and down-regulated DEGs for visualization:

```{r}
de_results$sign = sign(de_results$beta_Traj)
de_results = de_results[order(de_results$sign, de_results$p_Traj), ]

# omit MT- and other non-coding DEGs for visualization
de_results = de_results[!grepl("^(MT-|RP)", de_results$gene_ID), ]

# Identify the top up- and down-regulated DEGs for visualization
DEG_traj_overall = c(de_results$gene_ID[de_results$sign > 0][1:20],
            de_results$gene_ID[de_results$sign < 0][20:1])
```

Heatmap visualization for this cell type:

```{r message = FALSE, results = 'hide', warning = FALSE, fig.width=9, fig.height=7}
# Heatmap visualization for traj_overall (donors are ordered based on traj_overall)
sub_bulk_obj = ScaleData(sub_bulk_obj, features = DEG_traj_overall)
Scillus::plot_heatmap(dataset = sub_bulk_obj,
                      markers = DEG_traj_overall,
                      sort_var = "traj_overall",
                      anno_var = c("Disease"),
                      anno_colors = list(c("#f8766d", "cornsilk" )),
                      hm_colors = c("blue","white","red"))
```


